Computing in the Field of Rationals

### 2012-06-26 rational_ext.tex
_

\paragraph*{No division by zero.} 
The important side-condition on the types not captured by the syntax
above is that there are no types of the form {{1/b}}, where
{{[^b^]=0}} (see Def. \ref{def:size} and \ref{def:size-frac}
below). This eliminates terms of the form {{1/0}}, {{1/(0+0)}},
{{1/(0*b)}} etc, from being treated as valid {{b}} types. Intuitively,
this translates into the familiar restriction that there is no
division by 0 in arithmetic or the definition of fields in abstract
algebra.

@@

The rule {{(eta*)}} is different from other rules encountered before
due to the {{forall v:b}} quantification and is to be understood as
follows: when the machine encounters an {{eta*}} reduction it
non-deterministically replicates into several machine states or
worlds, each of which is seeded by a unique value {{v}} that inhabits
the type {{b}} that indexes {{eta* : 1 <-> 1/b * b}}. Each world has
an independant existence. As an example at the type {{bool}}, we would
have two possible machine states:

@@

The rule {{(eps*)}} is straightforward : when given a tuple
{{(1/v,v):1/b * b}} where {{eta*:1/b*b <-> 1}} the result is
{{()}}. The machine is undefined for cases where where the input is
{{(1/v',v):1/b*b}} and {{v' /= v}}.

@@

In the discourse below we use {{c v |-->* v'}} as a shorthand for the
execution sequence {{<c,v,[]>|-->*{[c,v',[] ]} }}.

@@

The above definition implies that combinators {{c:b1<->b2}} can be
indetified by relations between the sets {{b1}} and {{b2}}, i.e. 
{{c:b1<->b2 = {| (v, v') | v:b2, v':b2 and c v |-->* v'|} }}.
The definition of equality between combinators is extended reflect the
fact that combinators denote relations.

\begin{definition}[c1 = c2]
We say that {{c1 = c2}}, for combinators {{c1:b1<->b2}} and
{{c2:b1<->b2}} iff they denote the same relation between sets {{b1}}
and {{b2}}. Equivalently,

{{forall v:b1. eval(c1, v) = eval(c2, v)}}
\end{definition}

@@

Logical reversibility of {{PiEE*}} implies that {{c^{dagger} }}
denotes the inverse relation that {{c}} denotes, i.e.
{{c^^^={|(v',v)~|~(v,v') elem c|} }}.  Operationally, for any
execution path from {{v}} to {{v'}} in {{c}}, {{c{dagger} }} has an
execution path from {{v'}} to {{v}}.  Intuitively, logical
reversibility holds because for every {{eta*}} reduction in a world
that transforms {{()}} into {{(1/v,v)}}, on reverse execution the
adjoint {{eps*}} will transform {{(1/v,v)}} back to {{()}}.

@@

Like {{Pi}}, {{PiEE*}} is a commutative bimonoidal catgeory.
Symmetric monoidal categories where every object had a dual object and
where dualizing preserves the monodal tensor are referred to as
compact closed categories. Since divison is defined only for types of
non-zero size, not all objects have dual objects.  In this section
establish that {{PiEE*}} is ``locally compact closed'' in the sense
that for non-zero objects, namely types where {{[^ b ^]=0}}, it
satisfies the conditions for compact closed categories.

\begin{proposition}[Locally Compact Closed]
{{PiEE*}} is a locally compact closed category in the sense that for
all non-zero objects {{b}}, we have an inverse object {{1/b}} in the
symmetric monoidal structure of {{(*, 1)}}, with the morphisms:

@@

\noindent
The coherence conditions are more intuitive when presented
graphically. They require that the combinators below be equal to
{{id}}, at the appropriate type (see Sec. 4
\cite{springerlink:10.1007/978-3-642-12821-94}).

@@

  We briefly trace the execution of the first combinator (let us call
  it {{c}}. To show that {{c=id}}, we must show for {{v:b}}, the
  evaluation of {{<c,v,[]>}} always produces {{v:b}} as
  output. Verification for the second combinator is similar.

%subcode{opsem} include main
%! columnStyle = rllr
% & & <c, v:b, []>
% & |-->* & <eta*, (), C1>
% & |-->  & <eta*, (1/v',v'), C2> & some v':b
% (1) & |-->* & <eps*, (1/v', v), C3> & when v' = v
% & |-->~ & [eps*, (), C3]
% & |-->* & [c', v, [] ]
% (2) & |-->* & <eps*, (1/v', v), C3> & when v' /= v
% & |-->~ & undefined

@@

Without the important side condition on division by zero, we would
have {{1<->1/0*0}} which implies {{1 <-> 0}} and consequently
collapses the entire structure. Structures that are ``zero totalized
fields'' such as meadows \cite{Bergstra:2009:MES:1507774.1508000}
deserve investigation as alternative models for {{Pi}} with
fractionals (and negatives).

>>>>

\paragraph*{No division by zero.} 
The important side-condition on the types not captured by the syntax
above is that there are no types of the form {{1/b}}, where
{{[^b^]=0}} (see Def. \ref{def:size} and \ref{def:size-frac}
below). This eliminates terms of the form {{1/0}}, {{1/(0+0)}},
{{1/(0*b)}} etc, from being treated as valid {{b}} types. Intuitively,
this translates into the familiar restriction that there is no
division by 0 in arithmetic, nor in algebraic fields.

@@

The rule {{(eta*)}} is different from other rules encountered before
due to the {{forall v:b}} quantification. It is to be understood as
follows: when the machine encounters an {{eta*}} reduction it
non-deterministically replicates into several machine states (or
worlds). Each world is seeded by a unique value {{v:b}}, where {{b}}
comes from {{eta*:1 <-> 1/b * b}}.  For example, at
{{eta*:1<->1/bool*bool}}, we would have two possible machine
states:

@@

The rule {{(eps*)}} is straightforward : when given a tuple
{{(1/v,v):1/b * b}} where {{eps*:1/b*b <-> 1}} the result is
{{()}}. The machine is undefined for cases where where the input is
{{(1/v',v):1/b*b}} and {{v' /= v}}.

@@

\noindent
We use {{c v |-->* v'}} as a shorthand for the reduction sequence
{{<c,v,[]>|-->*{[c,v',[] ]} }}.

@@

The above definition implies that combinators {{c:b1<->b2}} denote
relations between the sets {{b1}} and {{b2}}, i.e.
{{c:b1<->b2={|(v,v') | v:b2, v':b2 and c v |-->* v'|} }}.  The
definition of combinator equality reflects the fact that combinators
denote relations.

\begin{definition}[c1 = c2]
We say that {{c1 = c2}}, for combinators {{c1:b1<->b2}} and
{{c2:b1<->b2}} iff they denote the same relation. Equivalently,
{{forall v:b1. eval(c1, v) = eval(c2, v)}}.
\end{definition}

@@

For any execution path from {{v}} to {{v'}} in {{c}}, {{c{dagger} }}
has an execution path from {{v'}} to {{v}}.  Stated differently,
logical reversibility implies that {{c^{dagger} }} denotes the inverse
relation that {{c}} denotes, i.e.  {{c^^^={|(v',v)~|~(v,v') elem c|} }}.  
Intuitively logical reversibility holds because for every
{{eta*}} reduction in a world that transforms {{()}} into {{(1/v,v)}},
on reverse execution the adjoint {{eps*}} will transform {{(1/v,v)}}
back to {{()}}.

@@

Like {{Pi}}, {{PiEE*}} is a commutative bimonoidal catgeory.
Symmetric monoidal categories where every object had a dual object and
where dualizing preserves the monoidal tensor are referred to as
compact closed categories. Since divison is defined only for types of
non-zero size, not all objects have duals.  We show that {{PiEE*}}
respects compact closed structure for non-zero objects. We introduce
the term ``locally compact closed'' to refer to this notion.

Without the important side condition on division by zero, we would
have {{1<->1/0*0}} which implies {{1 <-> 0}} consequently collapsing
the entire structure. In this context, an important area for future
investigation is to consider alternatives to fields as a model for
fractionals and negatives -- particularly algebraic structures that
are ``zero totalized fields'' such as meadows
\cite{Bergstra:2009:MES:1507774.1508000}.

\begin{proposition}[Locally Compact Closed]
For every non-zero object {{b}} of {{PiEE*}}, there exists an inverse
object {{1/b}} in the {{(*, 1)}} symmetric monoid with the morphisms:

@@

\noindent
The coherence conditions are intuitive when presented
graphically. They require that the combinators below be equal to
{{id}}, at the appropriate type (see Sec. 4
\cite{springerlink:10.1007/978-3-642-12821-94}).

@@

  We briefly trace the execution of the first combinator (let us call
  it {{c}}). To show that {{c=id}}, we must show for {{v:b}}, the
  evaluation of {{<c,v,[]>}} always produces {{v:b}} as
  output. Verification for the second combinator is similar.

%subcode{opsem} include main
%! columnStyle = rllr
% & & <c, v:b, []>
% & |-->* & <eta*, (), C1>
% & |-->  & <eta*, (1/v',v'), C2> & some v':b
% (1) & |-->* & <eps*, (1/v', v), C3> & when v' = v
% & |-->~ & [eps*, (), C3]
% & |-->* & [c, v, [] ]
% (2) & |-->* & <eps*, (1/v', v), C3> & when v' /= v
% & |-->~ & undefined

@@
